Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-03 paper.tex
Debloat, incorporate feedback.

\title{Decoupled-Fallback: A Portable Single-Pass GPU Scan}

@@

\begin{abstract}
  We present \emph{Decoupled-Fallback}, a method that enables single-pass \emph{Chained-Scans} to run on hardware without \emph{Forward-Progress Guarantees} while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory fences for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On an FPG-lacking device---Apple M1 Pro (8+2)---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-Then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by emulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-Then-Scan} even under extreme contention.
\end{abstract}

@@

\subsubsection{Inter-Workgroup Barriers}
Although threads \emph{within} a workgroup may synchronize with each other by participating in a \emph{barrier} primitive, the virtualization of processors introduces an emergent high-level limitation: the absence of an inter-workgroup barrier. As previously mentioned, workgroups cannot context switch. Although contemporary GPU hardware~\cite{} supports workgroup preemption for prioritizing real-time tasks or managing multi-process workloads, this functionality is deliberately excluded from the GPU programming model. This stems from the prohibitively high overhead in latency and storage associated with moving workgroup contexts on and off chip. In the oversubscription paradigm, where the number of workgroups is theoretically unbound and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups. Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points. However, kernel launches are suboptimal as barriers: they incur overheads due to the heterogeneous nature of the operation, and all execution context is lost between launches.

\subsubsection{Fairness and Progress}
Because multiple workgroups may reside on a single multiprocessor, two key issues arise at the subgroup scheduling level: \emph{fairness}---how evenly execution resources are distributed among threads---and \emph{progress guarantees}---ensuring that all threads or subgroups eventually make progress towards termination. However, the term "fairness" is used differently across the literature, leading to potential confusion. Sorensen et al.\cite{sorensen2016,sorensen2018,sorensen2021}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees. In contrast, NVIDIA publications~\cite{4523358,Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups. In this work, we adopt NVIDIA's terminology to maintain consistency with their publications, while acknowledging the broader definitions provided by Sorensen et al.

@@

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Contemporary GPU scans are distinguished from earlier PRAM-like scans~\cite{} by their alignment with the GPU memory hierarchy, which grades memory into progressively faster but increasingly private tiers. To operate efficiently, GPU scans must hybridize fine-grain intra-workgroup (\emph{local}) strategies, executed privately within shared memory and registers, with coarse-grain inter-workgroup (\emph{global}) strategies, performed in global memory. Due to their relatively low arithmetic intensity, it is theoretically expected and experimentally observed that the performance of local scans are memory-bound, as first demonstrated by Merrill and Grimshaw~\cite{}. Consequently, once equipped with a sufficient local scan strategy, the focus of scan design shifts to addressing the challenges of the global strategy, particularly inter-workgroup coordination and synchronization. In this section, we review the three prominent GPU global scan strategies---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---examining how they navigate inter-workgroup dependencies and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization.

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work tile partition of the input data. The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between work tiles; we call the result of this phase the \emph{root}. For large scans, this stage is an insignificant amount of work compared to the first and third.
  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and combines it with the result of the local scan to compute the final scan.
\end{enumerate}

As each work tile depends on the reductions of previous tiles, the absence of an inter-workgroup barrier necessitates kernel launches to resolve inter-workgroup dependencies. This results in the three-phase construction of \emph{Scan-then-Propagate}, with each phase implemented as a separate kernel launch. This is further complicated in Sengupta et al.'s implementation, where arbitrarily large inputs are processed through recursion. As the size of a work tile limits the maximum number of elements processed without inter-workgroup dependency resolution, the three-phase scan structure must be recursively applied until the spine size fits within a single work tile. Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches. Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total global data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

\subsubsection{Reduce-then-Scan}
If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic. This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{10.1145/1375527.1375559, Merrill2009, 10.1109/TPDS.2012.336, 10.5555/2031978.2032029}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. Like \emph{Scan-then-Propagate}, \emph{Reduce-then-Scan} employs a three-phase approach using kernel launches to resolve dependencies.
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
  \item \textbf{Spine Scan:} The spine is gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
\end{enumerate}

Merill and Grimshaw~\cite{Merrill2009} introduced a workgroup \emph{raking} technique that elides the overhead incurred when the recursive depth exceeds two. In this approach, the total workgroup occupancy $o$ is determined, and the input is partitioned into non overlapping blocks of size $\frac{n}{t \cdot o}$, which are raked serially by each workgroup. This limits the size of the spine from variably sized, proportional to the input size (possibly requiring recursion), to constant sized, and guaranteed to avoid further recursion. Assuming $t \geq o$, global memory movement during the spine scan is reduced from $O(n/t)$ to $O(2o)$. Kernel launches are limited to three and total global memory movement is reduced to $O(3n + 4o) = O(3n)$.

Breitbart~\cite{10.5555/2031978.2032029} was the first to apply more advanced inter-workgroup synchronization techniques to scan. Recognizing that workgroup raking naturally aligns with a \emph{Persistent-Threads}~\cite{gupta2012} paradigm, Breitbart utilized an atomics-based inter-workgroup barrier to replace kernel launches. Since multiprocessors are not oversubscribed in raking, workgroups remain resident (persist) on the multiprocessors for the entire lifetime of the kernel, enabling synchronization at the barrier without requiring context switching. Although global data movement remains at $O(3n)$, this consolidates the scan to a single kernel, minimizing launch overheads. \thomas{This also requires FPG, but not ideal to unpack and mention here . . .}

@@

In any \emph{Chained-Scan}, once a workgroup computes its local reduction, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. During this waiting period, the workgroup is continually scheduled, and the \emph{progress} of other workgroups must be guaranteed to prevent the spinning workgroup from starving its dependent predecessors of execution, which could result in deadlock.

@@

\subsubsection{Chained-Scan, Decoupled-Lookback}
Introduced by Merrill and Garland, the \emph{Decoupled-Lookback} technique separates the communication of reductions from dependency resolution, thus eliding bottlenecks caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.

@@

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time. Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers. However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is in exploring and formalizing progress models, rather than their application to any one algorithm.
\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct. The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA? It seems that it should be classified under OBE\@.}

The challenges faced by \emph{Chained-Scan} are analogous to the \emph{Dining Philosophers Problem}, where tasks compete for shared resources, potentially resulting in deadlock. Coffman et al.~\cite{10.1145/356586.356588} define four conditions necessary for deadlock:
\begin{enumerate}
  \item \textbf{Mutual Exclusion}: Tasks claim exclusive control of resources they require.
  \item \textbf{Resource Holding}: Tasks hold acquired resources while waiting.
  \item \textbf{No Preemption}: A resource cannot be removed from a task before completion.
  \item \textbf{Circular Dependency}:
        \begin{enumerate}
          \item Dependency Relationships: A dependency exists such that each task holds one or more resources that are required by proceeding tasks.
          \item Circular Ordering: A circular chain of tasks exists.
        \end{enumerate}
\end{enumerate}
Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1–3 are satisfied due to the one-to-one mapping of work tiles to workgroups and the constraints of the GPU programming model. Similarly, condition 4a is met by the dependency relationships between work tiles, leaving condition 4b---\emph{circular ordering}---as the key concern.

Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available before a workgroup begins its lookback. This uncertainty arises because workgroups are scheduled individually, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction \thomas{There is a better wording here. Something like, "because although workgroups \emph{may} share execution resources, they are ultimately scheduled individually." This sentence is incredibly important.}. In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available. During this waiting period, the spinning workgroup remains scheduled, consuming computational resources. Without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing the predecessor from completing its reduction. This creates a circular dependency, satisfying condition 4b and resulting in deadlock within the kernel.

For example, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor. Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@. In this setup, both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, loading the value at the memory address for the reduction of \emph{Work Tile 0} with each iteration. However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---its absence in key segments of hardware is a critical issue for developers. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\thomas{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery.} and at worst, risks the aforementioned TDR and subsequent program crash.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/Figure_1.pdf}
  \caption{Execution times of \emph{Decoupled-Lookback (PLACEHOLDER DATA)}}
\end{figure}

Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware. In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}. Yet, in the absence of guaranteed forward progress, developers have little choice but to sacrifice performance for broader compatibility.

\subsection{Earlier Attempts at Portability}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}. Each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time. By enabling workgroups to operate on data from other tiles, \emph{ScalarFallback} prevents deadlocking by targeting the \emph{mutual exclusion} condition described by Coffman et al. While this may involve performing redundant work when a tile is late rather than genuinely blocking, it ensures the algorithm's termination by allowing any workgroup to complete the reduction of a blocking tile, thereby breaking inter-workgroup dependencies.

\emph{ScalarFallback} suffers from two primary issues that impact its efficiency. First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin. This effectively takes $t$ spins to complete the fallback of a blocking $t$-sized tile. Moreover, because only a single thread is responsible for loading elements, the cache line associated with the loaded elements is severely underutilized, further degrading performance. Second, once the fallback is complete, the workgroup does not post its result to device memory. Consequently, every workgroup deadlocked by that tile must redundantly execute the fallback, wasting computational resources. While \emph{ScalarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan}.

\section{Decoupled-Fallback}
\emph{Decoupled-Fallback} separates fallback operations from the lookback process, enhancing the \emph{Scalar-Fallback} technique with significantly improved fallback performance. As an extension of the \emph{Chained-Scan} architecture, it retains the advantages of high speed and oversubscription, aligning with our design requirements.

Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to break deadlocking. \emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking. While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible. Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, leaving the same risks of deadlocking but at the multiprocessor level. While eliminating \emph{resource holding}---allowing workgroups to discard work tiles when they encounter a deadlock---is theoretically possible, it is strictly worse than eliminating \emph{mutual exclusion}, as it necessitates $t$ additional movements (discard-read-read versus read).
\begin{algorithm}
  \small
  \SetAlgoLined
  \KwGlobalIn{%
    $\textit{in}[]$,
    $\textit{out}[]$,
    $\textit{spine}[]$,
    $\textit{tileBump}$,
  }
  \KwSharedIn{%
    $\textit{wg\_partials}[]$,
    $\textit{wg\_fallback}[]$,
    $\textit{wg\_prev\_reduction}$,
    $\textit{wg\_broadcast}$,
    $\textit{wg\_control}$,
  }
  \KwGlobalOut{$\textit{out}[]$ inclusive scan of $\textit{in}[]$}
  \KwConstants{Vectors per thread $\textit{p}$}

@@

  \tcp{Decoupled Fallback}
  \While{$\textit{wg\_control} = \textnormal{LOCKED}$}{
    \textbf{Lookback}($\textit{spine}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
    \textbf{barrier()}\;

@@

Since WebGPU limits device-side timestamps to compute or render passes, we implement a custom Memcopy kernel for timing purposes. Similarly, as there are no existing WebGPU implementations of RTS, we rely on our own optimized version for comparison. To ensure a competitive evaluation, our RTS implementation employs a variable-launch main kernel with a single workgroup performing a serial spine scan, as we found raking to be less performant in practice. Apart from this adjustment, the intraworkgroup implementation in RTS is identical to that used in \emph{Decoupled-Fallback}.

@@

\section{Cut material, can be readded if we have enough space}

To ensure consistency with our coding artifact and promote cross-vendor compatibility, we use the terminology of the WebGPU specification.
\setlength{\tabcolsep}{3pt}
\begin{table}[h]
  \centering
  \caption{Non-Exhaustive Equivalent Terminology}
  \label{tab:terminology}
  \begin{tabular}{lcccc}
    \toprule
    \textbf{WGSL} & \textbf{GLSL} & \textbf{HLSL} & \textbf{Metal} & \textbf{CUDA} \\ \midrule
    Subgroup      & Subgroup      & Wave          & SIMD-group     & Warp          \\
    Workgroup     & Workgroup     & Group         & Threadgroup    & Block         \\
    Workgroup     & Shared        & Groupshared   & Threadgroup    & Shared        \\ \bottomrule
  \end{tabular}
\end{table}

>>>>

\title{Decoupled Fallback: A Portable Single-Pass GPU Scan}

@@

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained-Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory fences for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On an FPG-lacking device---Apple M1 Pro (8+2)---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by emulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\subsubsection{Inter-Workgroup Barriers}
The virtualization of processors introduces an emergent high-level limitation: the absence of an inter-workgroup barrier. As previously mentioned, workgroups cannot context switch. Although contemporary GPU hardware~\cite{} supports workgroup preemption for prioritizing real-time tasks or managing multi-process workloads, this functionality is deliberately excluded from the GPU programming model. This stems from the prohibitively high overhead in latency and storage associated with moving workgroup contexts on and off chip. In the oversubscription paradigm, where the number of workgroups is theoretically unbound and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups. Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points. However, kernel launches are suboptimal as barriers: they incur overheads due to the heterogeneous nature of the operation, and all execution context is lost between launches.

\subsubsection{Fairness and Progress}
Because multiple workgroups may reside on a single multiprocessor, two key issues arise at the subgroup scheduling level: \emph{fairness}---how evenly execution resources are distributed among subgroups---and \emph{progress guarantees}---ensuring that subgroups eventually make progress towards termination. However, the term "fairness" is used differently across the literature, leading to potential confusion. Sorensen et al.\cite{sorensen2016,sorensen2018,sorensen2021}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees. In contrast, NVIDIA publications~\cite{4523358,Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups. In this work, we adopt NVIDIA's terminology.

@@

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Contemporary GPU scans depart from early PRAM-like scans~\cite{} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers. Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory. Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{}. Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization. This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization. \thomas{I have written this paragraph like twenty times now, and it's still odious.}

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} extends intra-workgroup scans to large inputs by resolving inter-workgroup dependencies in three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:}  Each workgroup performs an inclusive scan on its assigned work tile, storing local results and reductions in global memory.
  \item \textbf{Spine Scan:} The workgroup reductions, collectively forming the \emph{spine}, are gathered and scanned to compute the \emph{root}, resolving dependencies between work tiles. For large scans, this stage is an insignificant amount of work compared to the first and third.
  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and adds it to the local scan output.
\end{enumerate}

The absence of inter-workgroup barriers necessitates separate kernel launches for dependency resolution, inherently structuring \emph{Scan-then-Propagate} into three phases. Sengupta et al.'s implementation further complicates this by using recursion to handle arbitrarily large inputs. As the size of a work tile limits the maximum number of elements processed without inter-workgroup dependency resolution, the three-phase structure must be recursively applied until the spine size fits within a single work tile. Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches. Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total global data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

\subsubsection{Reduce-then-Scan}
If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic. This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{10.1145/1375527.1375559, Merrill2009, 10.1109/TPDS.2012.336, 10.5555/2031978.2032029}: by storing only the work tile reduction in global memory during the first phase and performing a redundant scan in the third, it eliminates $n$ global data movement while preserving the three-phase structure:
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
  \item \textbf{Spine Scan:} The spine is scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
\end{enumerate}

Merrill and Grimshaw~\cite{Merrill2009} introduced \emph{workgroup raking} to further reduce recursion overhead when the recursion depth exceeds two. Instead of recursively reducing the spine, the input is partitioned into $o$ groups, where $o$ is the total workgroup occupancy. Each workgroup sequentially processes its assigned partition, reducing the spine size to a constant and guaranteeing a recursion depth of at most two. This reduces global memory movement for the spine scan from $O(n/t)$ to $O(2o)$ and limits kernel launches to three, reducing total global memory movement to $O(3n + 4o) = O(3n)$.

Breitbart~\cite{10.5555/2031978.2032029} applied Gupta et al.'s \emph{Persistent-Threads}~\cite{gupta2012} to scan, eliminating kernel launch overheads. Recognizing that raking naturally aligns with this paradigm, Breitbart replaced kernel launches with an atomics-based inter-workgroup barrier. Since multiprocessors are not oversubscribed in raking, workgroups persist for the entire kernel duration, allowing synchronization at the barrier without context switching. Despite maintaining $O(3n)$ global memory movement, consolidating the scan into a single kernel minimizes launch overheads.

@@

In any \emph{Chained-Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. During this waiting period, the workgroup is continually scheduled, and the \emph{progress} of other workgroups must be guaranteed to prevent the spinning workgroup from starving its dependent predecessors of execution, which could result in deadlock.

@@

\subsubsection{Chained-Scan, Decoupled-Lookback}
Introduced by Merrill and Garland, the \emph{Decoupled-Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.

@@

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (from possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation. Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers, classifying \emph{Chained-Scan} as FPG-reliant. However, their work primarily focuses on formalizing progress models rather than their direct application to specific algorithms.
\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct. The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA? It seems that it should be classified under OBE\@.}

The challenges faced by \emph{Chained-Scan} resemble the \emph{Dining Philosophers Problem}, where competing tasks share limited resources, risking deadlock. Coffman et al.~\cite{10.1145/356586.356588} define four conditions necessary for deadlock:
\begin{enumerate}
  \item \textbf{Mutual Exclusion}: Tasks claim exclusive control of resources they require.
  \item \textbf{Resource Holding}: Tasks hold acquired resources while waiting.
  \item \textbf{No Preemption}: A resource cannot be removed from a task before completion.
  \item \textbf{Circular Dependency}:
        \begin{enumerate}
          \item Dependency Relationships: A dependency exists such that each task holds one or more resources that are required by proceeding tasks.
          \item Circular Ordering: A circular chain of tasks exists.
        \end{enumerate}
\end{enumerate}
Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1–3 hold due to the one-to-one mapping of work tiles to workgroups and the constraints of the GPU programming model. Condition 4a is satisfied by the inter-tile dependencies, leaving condition 4b---\emph{circular ordering}---as the key concern.

Although the serial assignment of work tiles ensures that predecessor tiles are scheduled or completed, it does not guarantee their reductions will be available when a workgroup begins \emph{Chained Propagation}. This uncertainty arises because workgroups are scheduled \emph{individually}, meaning a predecessor may not have had the opportunity to compute and post its reduction. In such cases, the dependent workgroup spins, consuming execution resources while waiting. Because workgroups may share execution resources by being co-resident on the same multiprocessor, without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing progress and creating a circular dependency, satisfying condition 4b and resulting in deadlock.

For example, consider a scan operation requiring two workgroups on a GPU with a single multiprocessor. Both workgroups are co-resident, but the multiprocessor has only one scheduling unit, which lacks FPG@. Both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, repeatedly querying the memory address for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing shows that, at best, running \emph{Chained-Scan} without FPG results in severe performance degradation, and at worst, risks TDR and program crashes.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/Figure_1.pdf}
  \caption{Execution times of \emph{Decoupled-Lookback (PLACEHOLDER DATA)}}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-Then-Scan}, sacrificing performance for broader compatibility.

\subsection{Earlier Attempts at Portability}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{Decoupled-Lookback} approach without FPG requirements in \emph{Scalar Fallback}. Each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time. This eliminates the \emph{mutual exclusion} condition necessary for deadlock, as workgroups can operate on data from other tiles. While this may introduce redundant work when a tile is late rather than truly blocking, it guarantees termination by allowing any workgroup to complete a blocking tile’s reduction, breaking inter-workgroup dependencies.

\emph{Scalar Fallback} suffers from two primary inefficiencies. First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup enters a waiting spin. This results in a $t$-sized blocking tile requiring $t$ spins to complete. Additionally, since only one thread loads elements, the associated cache line is severely underutilized. Second, once the fallback completes, the workgroup does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback. While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.

\section{Decoupled Fallback}
\emph{Decoupled Fallback} separates fallback operations from the lookback process, enhancing the \emph{Scalar-Fallback} technique with significantly improved fallback performance. As an extension of the \emph{Chained-Scan} architecture, it retains the advantages of high speed and oversubscription, aligning with our design requirements.

Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to preclude deadlocks. \emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking. While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible. Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, shifting rather than eliminating deadlock risks. Removing \emph{resource holding}---allowing workgroups to discard \emph{blocked} work tiles---is theoretically possible but strictly inferior to removing \emph{mutual exclusion}, as it incurs $t$ extra memory operations (discard-read-read instead of read).
\begin{algorithm}
  \small
  \SetAlgoLined
  \KwGlobalIn{%
    $\textit{in}[]$,
    $\textit{out}[]$,
    $\textit{spine}[]$,
    $\textit{tileBump}$,
  }
  \KwSharedIn{%
    $\textit{wg\_partials}[]$,
    $\textit{wg\_fallback}[]$,
    $\textit{wg\_prev\_reduction}$,
    $\textit{wg\_broadcast}$,
    $\textit{wg\_control}$,
  }
  \KwGlobalOut{$\textit{out}[]$ inclusive scan of $\textit{in}[]$}
  \KwConstants{Vectors per thread $\textit{p}$}

@@

  \tcp{Chained Propagation, Decoupled Fallback}
  \While{$\textit{wg\_control} = \textnormal{LOCKED}$}{
    \textbf{Lookback}($\textit{spine}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
    \textbf{barrier()}\;

@@

@@

\begin{acks}
\end{acks}
